---
title: 'Assignment 10: Data Scraping'
author: "Sokna Kry"
output:
  pdf_document: default
  html_document:
    df_print: paged
geometry: margin=2.54cm
editor_options:
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on data scraping. 

## Directions
1. Rename this file `<FirstLast>_A10_DataScraping.Rmd` (replacing `<FirstLast>` with your first and last name).
2. Change "Student Name" on line 3 (above) with your name.
3. Work through the steps, **creating code and output** that fulfill each instruction.
4. Be sure your code is tidy; use line breaks to ensure your code fits in the knitted output.
5. Be sure to **answer the questions** in this assignment document.
6. When you have completed the assignment, **Knit** the text and code into a single PDF file.


## Set up 
1. Set up your session:

* Load the packages `tidyverse`, `rvest`, and any others you end up using.
* Check your working directory

```{r, message = FALSE}
#1
library(tidyverse)

library(rvest)
library(dbplyr)
library(lubridate)
library(ggplot2)

getwd()

# Set theme
mytheme <- theme_classic() +
  theme(axis.text = element_text(color = "black"), 
        legend.position = "top")
theme_set(mytheme)
```

2. We will be scraping data from the NC DEQs Local Water Supply Planning website, specifically the Durham's 2022 Municipal Local Water Supply Plan (LWSP): 
 * Navigate to https://www.ncwater.org/WUDC/app/LWSP/search.php
 * Scroll down and select the LWSP link next to Durham Municipality. 
 * Note the web address: <https://www.ncwater.org/WUDC/app/LWSP/report.php?pwsid=03-32-010&year=2022>
 
Indicate this website as the as the URL to be scraped. (In other words, read the contents into an `rvest` webpage object.)

```{r set.the.scraping.website}
#2
webpage <- read_html('https://www.ncwater.org/WUDC/app/LWSP/report.php?pwsid=03-32-010&year=2022')
webpage


the_registrant <- webpage %>% 
  html_nodes("div+ table tr:nth-child(1) td:nth-child(2)") %>% 
  html_text()
the_registrant

```

3. The data we want to collect are listed below:

* From the "1. System Information" section:
 * Water system name
 * PWSID
 * Ownership
 
* From the "3. Water Supply Sources" section:
 * Maximum Day Use (MGD) - for each month

In the code chunk below scrape these values, assigning them to four separate variables.

>HINT: The first value should be "Durham", the second "03-32-010", the third "Municipality", and the last should be a vector of 12 numeric values (represented as strings), with the first value being "36.1000".

```{r scrape.the.data}
#3
water.system.name <- webpage %>% 
  html_nodes("div+ table tr:nth-child(1) td:nth-child(2)") %>% 
  html_text()
water.system.name 
 
PWSID <- webpage %>% 
  html_nodes("td tr:nth-child(1) td:nth-child(5)") %>% 
  html_text()
PWSID

ownership <- webpage %>% 
  html_nodes("div+ table tr:nth-child(2) td:nth-child(4)") %>% 
  html_text()
ownership

max.withdrawals.mgd <- webpage %>% 
  html_nodes("th~ td+ td") %>% 
  html_text()
max.withdrawals.mgd

```


4. Convert your scraped data into a dataframe. This dataframe should have a column for each of the 4 variables scraped and a row for the month corresponding to the withdrawal data. Also add a Date column that includes your month and year in data format. (Feel free to add a Year column too, if you wish.)

>TIP: Use `rep()` to repeat a value when creating a dataframe.

>NOTE: It's likely you won't be able to scrape the monthly widthrawal data in chronological order. You can overcome this by creating a month column manually assigning values in the order the data are scraped: "Jan", "May", "Sept", "Feb", etc... Or, you could scrape month values from the web page...

5. Create a line plot of the average daily withdrawals across the months for 2022

```{r create.a.dataframe.from.scraped.data}
#4
df_max.withdrawals.mgd <- data.frame("Month" = rep(1:12),
                             "Year" = rep(2022,12),
                             "max.withdrawals_mgd" = as.numeric(max.withdrawals.mgd))

df_max.withdrawals.mgd <- data.frame("Month" = rep(1:12),
                             "Year" = rep(2022,12),
                             "max.withdrawals_mgd" = as.numeric(max.withdrawals.mgd))

df_max.withdrawals.mgd <- df_max.withdrawals.mgd %>% 
  mutate(water.system.name = !!water.system.name,
         PWSID = !!PWSID,
         ownership = !!ownership,
         Date = my(paste(Month,"-",Year)))

#5

#Plot 
ggplot(df_max.withdrawals.mgd,aes(x=Date,y=max.withdrawals.mgd, group=water.system.name)) + 
  geom_line() + 
  geom_smooth(method="loess",se=FALSE) +
  labs(title = paste("2022 Max Day Use of Water for",water.system.name),
       subtitle = water.system.name,
       y="Withdrawal (mgd)",
       x="Date")

```

6. Note that the PWSID and the year appear in the web address for the page we scraped. Construct a function using your code above that can scrape data for any PWSID and year for which the NC DEQ has data. **Be sure to modify the code to reflect the year and site (pwsid) scraped**.

```{r construct.a.scraping.function}
#6.

the_base_url <- 'https://www.ncwater.org/WUDC/app/LWSP/report.php?'
PWSID <- '03-32-010'
the_year <- 2015
the_scrape_url <- paste0(the_base_url, "pwsid=",PWSID, "&year=",the_year)
print(the_scrape_url)


scrape.it <- function(the_year, water.system.name)
  
  #Retrieve the website contents 
website <- read_html(paste0(the_base_url, "pwsid=",PWSID, "&year=",the_year))
website<- read_html(the_scrape_url)


  #Scrape the data items
water.system.name <- website %>% 
  html_nodes("div+ table tr:nth-child(1) td:nth-child(2)") %>% html_text()
PWSID <- website %>%   html_nodes("td tr:nth-child(1) td:nth-child(5)") %>% 
  html_text()
ownership <- website %>% 
  html_nodes("div+ table tr:nth-child(2) td:nth-child(4)") %>% html_text()
max.withdrawals <- website %>% html_nodes("th~ td+ td") %>% html_text()
  
  #Convert to a dataframe
df_withdrawals <- data.frame("Month" = rep(1:12),
                               "Year" = rep(the_year,12),
                               "max.withdrawals" = as.numeric(max.withdrawals)) %>% mutate(water.system.name = !!water.system.name,
        PWSID = !!PWSID,
        ownership = !!ownership,
        Date = my(paste(Month,"-",Year)))

  #Return the dataframe
return(df_withdrawals)

```

7. Use the function above to extract and plot max daily withdrawals for Durham (PWSID='03-32-010') for each month in 2015

```{r fetch.and.plot.Durham.2015.data}
#7

ggplot(df_withdrawals,aes(x=Date,y=max.withdrawals, group=water.system.name)) + 
  geom_line() + 
  geom_smooth(method="loess",se=FALSE) +
  labs(title = paste("2015 Max Day Use of Water for",water.system.name),
       subtitle = water.system.name,
       y="Withdrawal (mgd)",
       x="Date")
```

8. Use the function above to extract data for Asheville (PWSID = 01-11-010) in 2015. Combine this data with the Durham data collected above and create a plot that compares Asheville's to Durham's water withdrawals.

```{r fetch.and.plot.Asheville.2015.data}
#8

the_base_url <- 'https://www.ncwater.org/WUDC/app/LWSP/report.php?'
PWSID <- '01-11-010'
the_year <- 2015
the_scrape_url2 <- paste0(the_base_url, "pwsid=",PWSID, "&year=",the_year)
print(the_scrape_url2)


scrape.it <- function(the_year, water.system.name)
  
  #Retrieve the website contents 
  website <- read_html(paste0(the_base_url, "pwsid=",PWSID, "&year=",the_year))
  website<- read_html(the_scrape_url2)


  #Scrape the data items
  water.system.name <- website %>% 
    html_nodes("div+ table tr:nth-child(1) td:nth-child(2)") %>% html_text()
  PWSID <- website %>%   html_nodes("td tr:nth-child(1) td:nth-child(5)") %>% 
    html_text()
  ownership <- website %>% 
    html_nodes("div+ table tr:nth-child(2) td:nth-child(4)") %>% html_text()
  max.withdrawals <- website %>% html_nodes("th~ td+ td") %>% html_text()
  
  #Convert to a dataframe
  df_withdrawals <- data.frame("Month" = rep(1:12),
                               "Year" = rep(the_year,12),
                               "max.withdrawals" = as.numeric(max.withdrawals)) %>% mutate(water.system.name = !!water.system.name,
        PWSID = !!PWSID,
        ownership = !!ownership,
        Date = my(paste(Month,"-",Year)))
  
  
  #Return the dataframe
  return(df_withdrawals)


ggplot(df_withdrawals,aes(x=Date,y=max.withdrawals, group=water.system.name)) + 
  geom_line() + 
  geom_smooth(method="loess",se=FALSE) +
  labs(title = paste("2015 Max Day Use of Water for",water.system.name),
       subtitle = water.system.name,
       y="Withdrawal (mgd)",
       x="Date")


```


9. Use the code & function you created above to plot Asheville's max daily withdrawal by months for the years 2010 thru 2021.Add a smoothed line to the plot (method = 'loess').

> TIP: See Section 3.2 in the "09_Data_Scraping.Rmd" where we apply "map2()" to iteratively run a function over two inputs. Pipe the output of the map2() function to `bindrows()` to combine the dataframes into a single one. 

```{r fetch.and.plot.Asheville.multiyear.data }
#9 

#Code & function from Q#8
the_base_url <- 'https://www.ncwater.org/WUDC/app/LWSP/report.php?'
PWSID <- '01-11-010'
the_year <- 2015
the_scrape_url2 <- paste0(the_base_url, "pwsid=",PWSID, "&year=",the_year)
print(the_scrape_url2)


scrape.it <- function(the_year, water.system.name){
  
  #Retrieve the website contents 
  website <- read_html(paste0(the_base_url, "pwsid=",PWSID, "&year=",the_year))
  website<- read_html(the_scrape_url2)


  #Scrape the data items
  water.system.name <- website %>% 
    html_nodes("div+ table tr:nth-child(1) td:nth-child(2)") %>% html_text()
  PWSID <- website %>%   html_nodes("td tr:nth-child(1) td:nth-child(5)") %>% 
    html_text()
  ownership <- website %>% 
    html_nodes("div+ table tr:nth-child(2) td:nth-child(4)") %>% html_text()
  max.withdrawals <- website %>% html_nodes("th~ td+ td") %>% html_text()
  
  #Convert to a dataframe
  df_withdrawals <- data.frame("Month" = rep(1:12),
                               "Year" = rep(the_year,12),
                               "max.withdrawals" = as.numeric(max.withdrawals)) %>% mutate(water.system.name = !!water.system.name,
        PWSID = !!PWSID,
        ownership = !!ownership,
        Date = my(paste(Month,"-",Year)))

  #Return the dataframe
  return(df_withdrawals)
}

#Create dataframe

df_years<- seq(2010,2021) %>%
  map(scrape.it) %>%
  bind_rows()

ggplot(df_years,aes(x=Date,y=max.withdrawals,color=the_year)) + 
  #geom_line() + 
  geom_smooth(method="loess",se=FALSE) +
  labs(title = "Water usage data",
       y="Withdrawal (mgd)",
       x="Date")
```

> Question: Just by looking at the plot (i.e. not running statistics), does Asheville have a trend in water usage over time?
Yes. Asheville has a trend in water usage over time. The water usage is decreasing over time. 
